Forecasting the US Presidential Election: A Poll-of-Polls Approach Using Linear Models By Using Data of Pollsters From Five Thirty Eight In 2024

### 2024-11-03 paper/paper.qmd
add bayesian model for REP and predict results

```{=tex}
\begin{align} 
\log \left( \frac{p_i}{1 - p_i} \right) &= \beta_0 + \beta_1 \log(\text{sample\_size}_i) + \beta_2 \, \text{pollscore}_i + \beta_3 \, \text{numeric\_grade}_i \\
&+ \beta_4 \, \text{transparency\_score}_i + \beta_5 \, \text{days\_since\_start}_i + \alpha_j\\
\alpha_j &\sim \text{Normal}(0, \sigma_{\text{pollster}}) \\
\beta_0 &\sim \text{Normal}(0, 2.5) \\
\beta_1 &\sim \text{Normal}(0, 2.5) \\
\beta_2 &\sim \text{Normal}(0, 2.5) \\
\beta_3 &\sim \text{Normal}(0, 2.5) \\
\beta_4 &\sim \text{Normal}(0, 2.5) \\
\beta_5 &\sim \text{Normal}(0, 2.5) \\
\sigma_{\text{pollster}} &\sim \text{Exponential}(1)
\end{align}
```
$y_i$ represent the number of individuals in the sample that support the Democratic party (this corresponds to num_party). The response is modeled as binomial: $y_i \sim \text{Binomial}(\text{sample\_size}_i, p_i)$, where $p_i$ is the probability that an individual in poll $i$ supports the Democratic party, and $\text{sample\_size}_i$ is the total number of individuals surveyed in poll $i$.

Summary of Bayesian model is hsown in [@tbl-baymodelresults], and the ppcheck is in appendix [@fig-ppcheck].

@@

Bay_model <-
  readRDS(file = here::here("models/bayesian_model_dem.rds"))
```

@@

modelsummary::modelsummary(
  Bay_model,
  fmt = 2
)
```

# Prediction {#sec-predict}

To get a prediction by the Bayesian model we have, we first spline fit the model, and then create a data frame and generate posterior predictions by the new data frame.

## Spline Fit for Bayesian model set-up

We run the model in R [@citeR] using the `rstanarm` package of [@rstanarm]. We use priors $\text{Normal}(0, 5)$ to allow for more flexibility of predictors effects with “autoscale = TRUE”. Then we crate a data frame like [@tbl-newdata], we set the quality scale of predict data to the best to perform a poll that has high quality. the result is shown as [@sec-splresult]

@@

# Create new data for prediction
new_data <- data.frame(
  pollscore = -1.5,
  sample_size = 1200,
  numeric_grade = 3,
  transparency_score=10,
  days_since_start = seq(
    min(dem_data$days_since_start),
    max(dem_data$days_since_start),
    length.out = 100
  ),
  pollster = factor("TIPP", levels = levels(dem_data$pollster))
)

@@

@@

```{r}
#| echo: false
#| label: fig-splpredict
#| fig-cap: "Poll Percentage over Time with Spline Fit for DEM and REP party"
#| warning: false

@@

# Plot the spline fit
ggplot(combined_data, aes(x = days_since_start, y = pct, color = pollster)) +
  # DEM predictions
  geom_line(
    data = pred_summary %>% filter(party == "DEM"),
    aes(x = days_since_start, y = pred_mean_dem, color = "DEM"),
    inherit.aes = FALSE
  ) +
  geom_ribbon(
    data = pred_summary %>% filter(party == "DEM"),
    aes(x = days_since_start, ymin = pred_lower_dem, ymax = pred_upper_dem, fill = "DEM"),
    alpha = 0.2,
    inherit.aes = FALSE
  ) +
  # REP predictions
  geom_line(
    data = pred_summary %>% filter(party == "REP"),
    aes(x = days_since_start, y = pred_mean_rep, color = "REP"),  # Move color inside aes()
    inherit.aes = FALSE
  ) +
  geom_ribbon(
    data = pred_summary %>% filter(party == "REP"),
    aes(x = days_since_start, ymin = pred_lower_rep, ymax = pred_upper_rep, fill = "REP"),  # Move fill inside aes()
    alpha = 0.2,
    inherit.aes = FALSE
  ) +
  # Use scale_color_manual and scale_fill_manual for colors and fills
  scale_color_manual(values = c("DEM" = "blue", "REP" = "red"), 
                     name = "Party",
                     labels = c("DEM (Blue)", "REP (Red)")) +
  scale_fill_manual(values = c("DEM" = "blue", "REP" = "red"), 
                    name = "Party",
                    labels = c("DEM (Blue)", "REP (Red)")) +
  labs(
    x = "Days since 2024",
    y = "Percentage",
  ) +
  theme_minimal()
```

The prediction graph [@fig-splpredict] contains prediction of spline fit model for two parties, blue line represents the DEM party's predicted percentage of support, and the red line represents the REP party's predicted percentage of support over time (measured in days since the start of 2024). We can tell that the predicted support of DEM party is taking the lead by aroud 2 percent until the last day of the dataset.

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheck
#| fig-cap: "Posterior prediction check for Bayesian model"

pp_check(Bay_model)

```

>>>>

```{=tex}
\begin{align} 
\log \left( \frac{p_i}{1 - p_i} \right) &= \beta_0 + \beta_1 \log(\text{sample\_size}_i) + \beta_2 \, \text{pollscore}_i + \beta_3 \, \text{numeric\_grade}_i \\
&+ \beta_4 \, \text{transparency\_score}_i + \beta_5 \, \text{days\_since\_start}_i + \alpha_j\\
\alpha_j &\sim \text{Normal}(0, \sigma_{\text{pollster}}) \\
\beta_0 &\sim \text{Normal}(0, 2.5) \\
\beta_1 &\sim \text{Normal}(0, 2.5) \\
\beta_2 &\sim \text{Normal}(0, 2.5) \\
\beta_3 &\sim \text{Normal}(0, 2.5) \\
\beta_4 &\sim \text{Normal}(0, 2.5) \\
\beta_5 &\sim \text{Normal}(0, 2.5) \\
\sigma_{\text{pollster}} &\sim \text{Exponential}(1)
\end{align}
```
$y_i$ represent the number of individuals in the sample that support the Democratic party (this corresponds to num_party). The response is modeled as binomial: $y_i \sim \text{Binomial}(\text{sample\_size}_i, p_i)$, where $p_i$ is the probability that an individual in poll $i$ supports the Democratic party, and $\text{sample\_size}_i$ is the total number of individuals surveyed in poll $i$. We also create a model for Republican to compare the trend in prediction result in the following section [@sec-bayresult].

### Result {#sec-bayresult}
To get a prediction by the Bayesian model, we create a data frame and generate posterior predictions by the new data frame, example data looks like [@tbl-newdata], we set the quality scale of predict data to the best to perform a poll that has high quality. Summary of Bayesian model is shown in [@tbl-baymodelresults], and the posterior predictive check is in appendix [@sec-model-ppcheck]. The predict result is shown in [@fig-baypredict].

@@

bayesian_model_dem <-
  readRDS(file = here::here("models/bayesian_model_dem.rds"))
bayesian_model_rep <-
  readRDS(file = here::here("models/bayesian_model_rep.rds"))
```

@@

modelsummary::modelsummary(
  bayesian_model_dem,
  fmt = 2
)
```

@@

# Create new data for prediction
new_data <- data.frame(
  num_party = 0,
  pollscore = -1.5,
  sample_size = 1200,
  numeric_grade = 3,
  transparency_score=10,
  days_since_start = seq(
    min(dem_data$days_since_start),
    max(dem_data$days_since_start),
    length.out = 100
  ),
  pollster = factor("YouGov", levels = levels(dem_data$pollster))
)

@@

set.seed(123) # set seed for reproduce of predict result
predict_counts_dem <- posterior_predict(bayesian_model_dem, newdata = new_data)
predict_counts_rep <- posterior_predict(bayesian_model_rep, newdata = new_data)

combined_data <- bind_rows(dem_data, rep_data)
pred_summary_dem <- new_data %>%
  mutate(
    pred_mean_dem = colMeans(predict_counts_dem),
    pred_lower_dem = apply(predict_counts_dem, 2, quantile, probs = 0.025),
    pred_upper_dem = apply(predict_counts_dem, 2, quantile, probs = 0.975)
  )

# Summarize predictions for REP
pred_summary_rep <- new_data %>%
  mutate(
    pred_mean_rep = colMeans(predict_counts_rep),
    pred_lower_rep = apply(predict_counts_rep, 2, quantile, probs = 0.025),
    pred_upper_rep = apply(predict_counts_rep, 2, quantile, probs = 0.975)
  )

# Combine both summaries
combined_summary <- pred_summary_dem %>%
  select(days_since_start, pred_mean_dem, pred_lower_dem, pred_upper_dem) %>%
  left_join(
    pred_summary_rep %>% 
      select(days_since_start, pred_mean_rep, pred_lower_rep, pred_upper_rep),
    by = "days_since_start"
  )
# Extract the last day data for annotations
last_day_data <- combined_summary %>%
  filter(days_since_start == max(days_since_start))
```

```{r}
#| echo: false
#| label: fig-baypredict
#| fig-cap: "Predicted number of supporters for DEM and REP party"
#| warning: false

ggplot(combined_summary, aes(x = days_since_start)) +
  geom_line(aes(y = pred_mean_dem, color = "DEM"), size = 1) +
  geom_ribbon(aes(ymin = pred_lower_dem, ymax = pred_upper_dem, fill = "DEM"), alpha = 0.2) +
  geom_line(aes(y = pred_mean_rep, color = "REP"), size = 1) +
  geom_ribbon(aes(ymin = pred_lower_rep, ymax = pred_upper_rep, fill = "REP"), alpha = 0.2) +
  scale_color_manual(values = c("DEM" = "blue", "REP" = "red")) +
  scale_fill_manual(values = c("DEM" = "blue", "REP" = "red")) +
  geom_text(
    data = last_day_data,
    aes(
      x = days_since_start,
      y = pred_mean_dem,
      label = paste("DEM:", round(pred_mean_dem))
    ),
    color = "blue",
    vjust = -1,
    hjust = 1
  ) +
  geom_text(
    data = last_day_data,
    aes(
      x = days_since_start,
      y = pred_mean_rep,
      label = paste("REP:", round(pred_mean_rep))
    ),
    color = "red",
    vjust = -1,
    hjust = 1
  ) +
  labs(
    x = "Days Since Start",
    y = "Predicted Number of Supporters",
    color = "Party",
    fill = "Party"
  ) +
  theme_minimal()
```

The prediction results of the Bayesian model closely align with those of the linear model, with both showing the mean predicted number of supporters for each party. Notably, the Democratic Party takes the lead around day 210 (late July). By early November, the predictions indicate that the Democratic Party has an average of 601 supporters, compared to the Republican Party's 578. This suggests a slight advantage for the Democrats as the final prediction period concludes.

## Spline Fit model set-up

We first run the model in R [@citeR] using the `rstanarm` package of [@rstanarm], and use priors $\text{Normal}(0, 5)$ to allow for more flexibility of predictors effects with “autoscale = TRUE”. Then we use the same data frame that used for predict for Bayesian model [@tbl-newdata], the result is shown as [@sec-splresult]

@@

@@

# Extract the last day data for annotations
last_day_data <- pred_summary %>%
  filter(days_since_start == max(days_since_start))
```

```{r}
#| echo: false
#| label: fig-splpredict
#| fig-cap: "Predicted support percentage over Time with Spline Fit for DEM and REP party"
#| warning: false

ggplot(pred_summary, aes(x = days_since_start)) +
  geom_line(aes(y = pred_mean_dem, color = "DEM"), size = 1, data = subset(pred_summary, party == "DEM")) +
  geom_ribbon(aes(ymin = pred_lower_dem, ymax = pred_upper_dem, fill = "DEM"), alpha = 0.2, data = subset(pred_summary, party == "DEM")) +
  geom_line(aes(y = pred_mean_rep, color = "REP"), size = 1, data = subset(pred_summary, party == "REP")) +
  geom_ribbon(aes(ymin = pred_lower_rep, ymax = pred_upper_rep, fill = "REP"), alpha = 0.2, data = subset(pred_summary, party == "REP")) +
  scale_color_manual(values = c("DEM" = "blue", "REP" = "red")) +
  scale_fill_manual(values = c("DEM" = "blue", "REP" = "red")) +
    geom_text(
    data = last_day_data,
    aes(
      x = days_since_start,
      y = pred_mean_dem,
      label = paste("DEM:", round(pred_mean_dem, 2))
    ),
    color = "blue",
    vjust = -1,
    hjust = 1
  ) +
  geom_text(
    data = last_day_data,
    aes(
      x = days_since_start,
      y = pred_mean_rep,
      label = paste("REP:", round(pred_mean_rep, 2))
    ),
    color = "red",
    vjust = -1,
    hjust = 1
  ) +
  labs(
    x = "Days Since Start",
    y = "Predicted Number of Supporters",
    color = "Party",
    fill = "Party"
  ) +
  theme_minimal()
```

The prediction graph [@fig-splpredict] contains prediction of spline fit model for two parties, blue line represents the DEM party's predicted percentage of support, and the red line represents the REP party's predicted percentage of support over time (measured in days since the start of 2024). We can tell that the predicted support of DEM party is taking the lead by around 2 percent until the last day of the dataset.

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheck
#| fig-cap: "Posterior prediction check for Bayesian models"
#| layout-ncol: 2
#| fig-subcap: ["Democratic Model", "Republican Model"]

pp_check(bayesian_model_dem)
pp_check(bayesian_model_rep)
```
